\section*{Acknowledgements}

%% Make the paragraph spacing different for this page. (Undone at the bottom)
\newlength\oldparskip
\setlength{\oldparskip}{\parskip}
\setlength{\parskip}{1cm}


This thesis and research work was a significant undertaking and would not have been possible without the love and support of others.

First and foremost, I wish to thank my supervisor Jonathan Keith who gave me this incredible opportunity with such an interesting project; thank you for your constant support, your invaluable mentorship, your patience and kindness. Thank you for taking the leap.

I am also deeply grateful to my associate advisor Kerrie Mengersen, for introducing me to researchers and fellow students who share similar interest with me and who gave me valuable advice on my research. Thank you for broadcasting and celebrating my successes and for your continuous support.

Thank you to Danny Spring, whose emails were always an eye opener and who gave me a glimpse on the many open problems in ecology. You have helped me focusing on what is relevant and important and pointed me into the right directions many times. 

Also, thank you to Tim Garoni, who introduced me to ACEMS and was always supportive after my presentations and talks which were draining experiences especially at the beginning.

A big thank you goes to ACEMS that organised many interesting talks, seminars, retreats and workshops. In particular, I really want to thank Louise Ryan, that took time to check on me during the Covid-19 pandemic and the lockdown.

Thank you to Mark Flegg, for asking the right questions and to Andrea Collevecchio, for his honest feedback and positive attitude.

I am also grateful for the administrative support provided by the School of Mathematics at Monash University. In particular, I wish to thank John Chan for his help in organising presentations, milestones, and conference attendances. Thank you for keeping the post-graduate students together and thank you for listening.

Finally, thank you to my family, always on my side. To my kids, for being so good and easy and to my husband, who supported me through what seemed like an impossible journey, and saw it to the end with me. Without your support this would not have been possible. Without your love I would not have been able to sustain the emotional roller-coaster that a research project is.
Thank you mum for teaching me the importance of resilience and hard work.
Thank you dad for instilling in me the love for science.


%{\noindent
%Thanks.
%}
%
%{\noindent
%More thanks maybe.
%}


\setlength{\parskip}{\oldparskip}
