\chapter[A novel Spatio-temporal self-exciting point process to model invasive species]{a novel Spatio-temporal self-exciting point process to model invasive speciess}
\label{ch:Model}

Here we introduce a new model for the RIFA invasion based on a self-exciting point process.

In this new approach we are seeking to simplify and improve on Keith and Spring agent-based model \cite{Keith} with the aim to reduce the running time without compromising on flexibility and completeness.

The model of Keith and Spring focused on reconstructing the historical trajectory of the invasion to determine if the current eradication strategy was successful. Their method consisted of constructing a likelihood model in terms of some unknown parameters that included, among other things, the phylogeny, jump type, founding type and treatment success rate. They then constructed the dependencies among the known and unknown parameters and defined all conditional distributions.

The posterior distribution was then sampled using a generalised Gibbs technique that enables transdimensional sampling, which is used when the number of parameters is unknown, as in their case.

A key feature of our model is that it will not need to include the phylogeny of the nests in the likelihood, which will result in a much lower computational time when running the inference code.

This will also allow the model to be adapted to other type of invasions where control strategies will have to be decided rapidly as new data is acquired.

We have been also considering the detection processes in parallel with the founding events in a similar way as in Jewell's model for infectious diseases \cite{Jewell}.

We then included unobserved nests alongside observed nests in the detection likelihood to fully model the missing data.

Finally, we have considered events as happening continuously in time.

\section{RIFA Model}

Let us consider a self-exciting spatial-temporal point process $N$ as a generalization of an Hawkes model. We can then specify an intensity function $\lambda(x, y, t)$ which represents the infinitesimal expected  rate of events at time $t$ and location $(x, y)$  given all the events up to time $t$.

Here $\mathcal{F}$ is a $\sigma$-algebra on $S \times [0, \infty )$, where $S$ is a bounded region of $\mathbb{R}$ and $[0, \infty)$ is the time interval. $N: \mathcal{F} \to \mathbb{R}$ and $N(A)$ is the number of points in $A$ for $A \in \mathcal{F}$

We will consider $N$ to be an inhomogeneous Poisson process, for which $\lambda(x, y, t)$ depends only on $(x, y)$ and $t$ \cite{Schoenberg}. The function $g(x - x', y - y', f - f')$ is the clustering density, where $f'$ are the times of founding of parents nests and the spatial coordinates $(x', y')$ will specify the location $s'$ of a parent nest.


We can then write the conditional intensity function for new nests considering the full past history in term of a stocastic integral:

\[
\lambda(x, y, t) = \mu(x, y, t) + \int_{0}^{t} \iint_{S} g(x - x', y - y', f - f') \d N(x', y', f')
\]
$\d N(x', y', f') = 1$ if the infinitesimal element $\d N(x', y', f')$ contains a parent and is 0 otherwise. We consider the temporal behavior of the process as independent of the spatial behavior so we can write

\[
\lambda(x, y, t) = \mu(x, y, t) + \int_{0}^{t} \iint_{S} m(f-f') \cdot l((x - x'), (y - y')) \d N(x', y', f')
\]
Where $\mu$ is a background term, while $m$ and $l$ are the triggering functions for time and space respectively. 

New individuals from the invasive species can be introduced at any time by carriers like animals or humans. For simplicity, in the RIFA invasion we will assume that these exogenous introductions are not happening. Also we assume to know time and location of the first nest.  We will therefore set the background term $\mu(x, y, f) $ equal to 0.

Each parent nest will be able to found more than one nest, with the number of nests founded per nest per month being a parameter $\zeta$, therefore the temporal triggering kernel will be a step function. Also the nest will have a maturation time $t_m$ of 8 months. Before this time they will not be able to produce new nests. 

The step function will be:


\[
m (f - f') =
\begin{cases}
0, & \mbox{if} \quad f - f' < t_{m} \\
\zeta, & \mbox{if} \quad f - f' \geq t_{m}
\end{cases}
\]
Where $f$ is the founding time of the new nest and $f'$ is the founding time of the parent nest.

For the newly founded nests we will consider a radial distance from the parent nest with an exponential distribution. The distribution over the angular direction will be uniform as it is equally possible to found a nest in every direction from the parent nest.

\begin{equation}
l(x - x', y - y')= J \bigg(\frac{1}{2 \pi} \sigma e^{- \sigma r}\bigg)
\end{equation}
where $J$ is the Jacobian

\[
J =  
\begin{vmatrix}
	\frac{\partial r}{\partial x} & \frac{\partial r}{\partial y} \\
	\frac{\partial \theta}{\partial x} & \frac{\partial \theta}{\partial y} \\
\end{vmatrix}
\]
where

\[
r^{2} = (x - x')^{2} + (y - y')^{2}
\]
and

\[
\theta = tan^{-1} \Bigg [\frac{y - y'}{x - x'} \Bigg ]
\]
therefore

\[
J = \frac{1}{\sqrt{(x - x')^{2} + (y - y')^{2}}}.
\]
We also make the assumption that nests are killed as soon as they are detected, so we introduce an indicator function $I(t' - t)$ such that

\[
I (t' - t) =
\begin{cases}
1, & \mbox{if} \quad t' -  t> 0 \\
0, & \mbox{otherwise}
\end{cases}
\]
where $t'$ represents the time of detection. So the conditional intensity function will be

\[
\lambda(x, y, t) = \int_{0}^{t} \int_{0}^{t} \iint_{S} m(f - f') \cdot I(t' - t)\cdot \frac{\sigma J }{2 \pi} e^{- \sigma r} \d N(x',y',t',f')
\]
where now the term $\d N(x',y',t',f')$ is 1 if a nest was founded in an infinitesimal time about $f'$ at a location in an infinitesimal region about $(x', y')$ and was detected in an infinitesimal time around $t'$ and is $0$ otherwise. Note that we must have $\d N(x',y',t',f')=0$ whenever $f' \geq t'$.

The likelihood at time $T$ will then be that of  an inhomogeneous Poisson process for the founding process with intensity $\lambda(x, y, t)$.

Also, the discovery time $t$ of a nest is known, but it is unknown when the nest was founded. So the time from establishment to notification $(t - f)$ of a nest $i$ is a random variable which we will consider exponentially distributed.

\[
h(t_{i} - f_{i}) = \gamma \exp (- \gamma(t_{i} - f_{i}))
\]
The Likelihood for a set of $n$ locations $(s_{1}, ... , s_{n})$, a set of founding times $f_{1}, ... , f_{n}$, and detection times $(t_{1},  ... , t_{n})$ will then be:

\[
\begin{aligned}
L(s_{1}, ..., s_{n}, f_{1}, ..., f_{n}, t_{1}, ..., t_{n} | \Theta) = & \Bigg[ \prod_{j=1}^{n} \lambda(s_{j}, t_{j}) \Bigg] \times \exp \Bigg(- \int_{0}^{T} \int_{S} \lambda(s, t) \d s \d t \Bigg) \times \\ 
& \times \prod_{\{ i : t_{i} < T \} } h (t_{i} - f_{i}) \times \prod_{ \{ i : t_{i} = \infty \} } \int_{T}^{\infty} h(t - f_{i}) \d t
\end{aligned}
\]
where $t_{i} = \infty$ if nest $i$ has not been detected at time $T$. $h(t_{i} - f_{i})$ is the contribution of the observed nests, while $\int_{T}^{\infty} h(t - f_{i}) dt$ is the contribution of the unobserved nest. $\Theta= ( \sigma, \gamma, \zeta)$ is the vector of the unknown parameters and $n$ is the number of nests founded.

The term 

\begin{equation}
\exp \bigg(- \int_{0}^{T} \int_{S} \lambda(s, t)\d s \d t \bigg)
\end{equation}
in the likelihood will evaluate to 


\[
\exp \bigg(- \zeta \sum_{i=1}^{n} (min\{ T, t_i \} - f_i) \bigg)
\]
as the spatial part evaluates to 1 and the temporal part is constant over the lifetime of each nest.

The likelihood can then be rewritten:

\[
\begin{aligned}
L(s_{1}, ..., s_{n}, f_{1}, ..., f_{n}, t_{1}, ..., t_{n} | \Theta) = & \Bigg[ \prod_{j=1}^{n} \lambda(s_{j},f_{j}, t_{j}) \Bigg] \times \exp \bigg(-\zeta \sum_{i=1}^{n} (min\{ T, t_i \} - f_i) \bigg)  \times \\
& \times \prod_{\{ i : t_{i} < T \} }  h (t_{i} - f_{i}) \times \prod_{ \{ i : t_{i} = \infty \} } \int_{T}^{\infty} h(t - f_{i}) \d t
\end{aligned}
\]
And the log-likelihood:

\[
\begin{aligned}
l(s_{1}, ..., s_{n}, f_{1}, ..., f_{n}, t_{1}, ..., t_{n} | \Theta) = & \Bigg[ \sum_{j=1}^{n} \log \lambda(s_{j},f_{j}, t_{j}) \Bigg] - \bigg(\zeta \sum_{i=1}^{n} (min\{ T, t_i \} - f_i) \bigg)  + \\
& + \sum_{\{ i : t_{i} < T \} }  \log h (t_{i} - f_{i}) + \sum_{ \{ i : t_{i} = \infty \} } \log \int_{T}^{\infty} h(t - f_{i}) \d t
\end{aligned}
\]
Substituting $h$ and evaluating the last integral we get 

\[
\begin{aligned}
l = & \Bigg[ \sum_{j=1}^{n} \log \lambda(s_{j},f_{j}, t_{j}) \Bigg] - \bigg(\zeta \sum_{i=1}^{n} (min\{ T, t_i \} - f_i) \bigg)  + \sum_{\{ i : t_{i} < T \} }  \bigg[\log (\gamma) -\gamma(t_{i} - f_{i}) \bigg] \\
+ & \sum_{ \{ i : t_{i} = \infty \} } \bigg[\log \bigg(\frac{1}{\gamma}\bigg) -\gamma(T - f_{i}) \bigg]
\end{aligned}
\]
In Bayesian statistics, we begin with prior distributions $P(\Theta)$ which encodes all that is known about $\Theta$ based on previous knowledge. After analyzing the data, the posterior distribution encodes all information about $\Theta$ based on both the prior and the data. In our case we want to keep the priors uninformative, therefore we choose gamma distributions with scale and shape equal to 1. The posterior distribution then is give by:

\[
P(\Theta | s_{1}, ..., s_{n}, f_{1}, ... , f_{n}, t_{1}, ..., t_{n}) \propto L(s_{1}, ..., s_{n}, f_{1}, ... , f_{n}, t_{1}, ..., t_{n} | \Theta) \times P(\Theta)
\]

\section{Model simulation} \label{section:simulationModel}

For the simulation of the process in a time interval $\tau_j = (T_{j-1}, T_j]$ between observations we will view the model as a branching Poisson process \cite{Lewis} which is a natural choice for invasive species. This immigration-birth representation gives rise to a simpler simulation procedure than the thinning algorithm introduced by Ogata \cite{Ogata}. A procedure of this kind was introduced by Zhuang \cite{Zhuang} for self-exciting spatial temporal point processes in the context of earthquake models. The simulation will proceed as follow:

\begin{enumerate}
    \item Consider the catalogue of events $G^{T_{j-1}}$ which contains the nests that have not been observed and killed at or before time $T_{j-1}$. Take all the unobserved nests that will be mature (older than 8 months) by time $T_j$ and put them in the catalogue $G_l^{T_j}$.
    \item Set $l = 0$.
    \item From $G_l^{T_j}$ select which nests are mature and put in catalogue $M_l^{T_j}$.
    \item For each nest in $M_l^{T_j}$ simulate the number of offspring $\nu^{T_j}$ produced in the time interval $\tau_j$ from a Poisson distribution with parameter $ m^{T_j} = \zeta (T_j-t_m)$ where $m^{T_j}$ was defined in equation (\ref{eq:NumOffsp}).
    \item Simulate for each of the offspring the time of founding from a uniform distribution that is $1/(T_j-t_m)$ if $t_m \leq f - f' \leq T_j$ and $0$ otherwise.
    \item Simulate the number of long-distance jumps $\mu^{T_j}$ from a binomial distribution with probability $\gamma$ and total number of offspring $\nu^{T_j}$.
    \item Simulate the location of these offspring. This is done simulating the angular direction and the radial direction separately, then calculating the coordinates $x$ and $y$. The angular direction will be simulated from a uniform distribution in the interval $(0, 2\pi]$. The radial direction for the long jumps will be simulated form a L\'evy distribution with parameter $c$ (VVV the bigger this parameter the higher the jumps simulated). The radial direction for the cluster events will be simulated from a decaying exponential distribution with parameter $\rho$.
    \item Simulate for each offspring the time from establishment to notification from an exponential distribution with parameter $\varphi$.
    \item Delete all parent nests that have been observed in the time interval $\tau_j$.
    \item Store each newly generated and not yet observed nest's location, time of founding and time of observation in $O_l^{T_j}$. Let $G_{l}^{T_j} = \bigcup_{k=1}^{\nu^{T_j}} (O_l^{T_j})_k$.
    \item If $G_l^{T_j}$ is not empty, set $l = l+1$ and return to step 3. Otherwise return $\bigcup_{h=0}^l G^{T_j}_h$ as the final set of simulated events for the time interval $\tau_j$.

\end{enumerate}

